### 2022-07-29 paper/localpsf.tex
more edits. Got to just before computational cost section
Paper draft is good other than some minor needed reference fixes. Ready for editing pass from Noemi

The ability to rapidly evaluate approximate kernel entries $\Aker(y,x)$ allows us to construct an H-matrix approximation, $\boldsymbol{\Aker}_H \approx \mathbf{\Aker}$, using the conventional adaptive cross H-matrix construction method (Section \ref{sec:h_matrix_construction_short}). In this method, one forms low rank approximations of off-diagonal blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. We then use H-matrix methods to convert $\boldsymbol{\Aker}_H$ into an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$. Schematically the algorithm proceeds as follows:
\begin{equation*}
\Aop \longrightarrow \spatialvol, \spatialmean, \spatialcov \longrightarrow \underbrace{\{x_i\}_{i=1}^r}_{\text{as }S_b\text{'s}} \longrightarrow \underbrace{\{\impulseresponse_{x_i}\}_{i=1}^r}_{\text{as }\eta_b\text{'s}} \longrightarrow \boldsymbol{\Aker}_H \longrightarrow \mathbf{A}_H.
\end{equation*}
The complete algorithm for constructing $\mathbf{A}_H$ is shown in Algorithm \ref{alg:construct_Atilde}. The cost is discussed in Section \ref{sec:overall_cost}.

@@

The computational cost of our method may be divided into the costs to perform the following tasks: (1) Computing impulse response moments and batches; (2) Building the H-matrix; (3) Performing linear algebra operations with the H-matrix. This may optionally include modifying the H-matrix to make it symmetric positive semi-definite.
In practice, (1) is the dominant cost, because (1) is the only task that requires applying $\Aop$ to vectors, and our method is targeted at applications in which applying $\Aop$ to a vector requires an expensive computational procedure such as solving a large-scale partial differential equation. All operations that do not require applications of $\Aop$ to vectors are nearly-linear, and therefore scalable, in the size of the problem, $N$.
We now describe these costs in detail. For convenience, Table \ref{tab:vars} lists variable symbols and their approximate sizes.

@@

\paragraph{(1) Computing impulse response moments and batches} Computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ requires applying $\Aop$ to $1$, $\gdim$, and $\gdim(\gdim+1)/2$ vectors, respectively.
Computing each impulse response batch requires applying $\Aop$ to one vector, so computing the impulse response batches $\{\eta_i\}_{i=1}^{\nbatch}$ requires $\nbatch$ operator applies. In total, computing the impulse response moments and batches therefore requires
\begin{equation*}
	1 + \gdim + \gdim(\gdim+1)/2 + \nbatch \qquad \text{operator applies.}
\end{equation*}
In a typical application one might have $\gdim=2$ and $\nbatch=5$, in which case a modest $11$ operator applies are required.

@@

\paragraph{(2) Building the H-matrix} The classical H-matrix construction technique
%, which is described in Appendix \ref{app:h_matrix}, 
requires evaluating $O(\hrank^2 \fedim \log \fedim)$ matrix entries of the approximation, where $\hrank$ is the H-matrix rank, i.e,  the maximum rank among the blocks of the H-matrix. To evaluate one matrix entry, first one must find the $\numnbr$ nearest sample points to a given point, where $\numnbr$ is the number of impulses used in the RBF interpolation. This is done using a precomputed kd-tree of sample points, and requires $O(\numnbr \log \nsamplepts)$ floating point and logical elementary operations. Second, one must find the mesh cells that the points $\{z_i\}_{i=1}^{\numnbr}$ reside. This is done using an AABB tree of mesh cells, and requires $O(\numnbr \log \fedim)$ elementary operations. Third, we must evaluate finite element basis functions on each of those cells, which requires $O(\numnbr)$ elementary operations. Finally, we must solve a $\numnbr \times \numnbr$ linear system to perform the radial basis function interpolation, which requires $O(\numnbr^3)$ elementary operations, which typically is negligible. Therefore, building the H-matrix requires
\begin{equation*}
	O\left(\left(\hrank^2 \fedim \log \fedim\right)\left(\numnbr \log \fedim + \numnbr^3\right)\right) \qquad \text{elementary operations}.
\end{equation*}
%This follows from the above discussion, and the fact that $\fedim \ge \nsamplepts$.

\paragraph{(3) Performing linear algebra operations with the H-matrix} It is well-known that H-matrix methods for matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion, and low rank updates require
\begin{equation}
\label{eq:H_matrix_cost_operations}
	O(\hrank^2 \fedim \log(\fedim)^a) \qquad \text{elementary operations},
\end{equation}
where $a \in \{0,1,2,3\}$ \cite{GrasedyckHackbusch03}. The precise cost depends on the type of H-matrix used and the operation being performed. In this paper, we use one matrix-matrix addition to add the H-matrix approximation of the data misfit term in the Hessian to the regularization term in the Hessian. Then, we use one matrix inversion to form the inverse of the Hessian, which is used as a preconditioner for the Newton linear system. 
%We perform low-rank updates to the H-matrix as the Newton iterations proceed, as per the discussion in Appendix \ref{sec:recycle_krylov}.
The rational method for modifying $\mathbf{A}_H$ to be positive semi-definite, described in Appendix \ref{sec:make_spd}, is performed using two matrix-matrix additions, $\ratord+1$ matrix-matrix multiplications, and one matrix inversion, where $2^\ratord$ is the rational function order.

@@

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to $\Gamma_b$. The basal boundary, $\Gamma_b$, is a 2D surface embedded in 3D due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate. The parameter $\basalfriction$ is dually viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building our Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space.

@@

The reconstructed parameter field $\basalfriction$ is found as the solution to an optimization problem of the following form:
\begin{equation}
	\label{eq:deterministic_optimization_problem}
	\min_{\basalfriction} \quad J(\basalfriction) := J_d(\basalfriction) + J_r(\basalfriction).
\end{equation}
The first term in the objective function is
\begin{equation*}
	J_d(\basalfriction) := \frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS.
\end{equation*}
We call this term the data misfit, because it measures the mismatch between the observed data and the predicted data based on a candidate parameter $\basalfriction$. Here $\velocity = \velocity(\basalfriction)$ denotes the velocity field solving $\ref{Stokeseqn}$ for the given parameter $\basalfriction$. The second term,
\begin{equation*}
	J_r(\basalfriction) := \frac{1}{2}\int_{\Omega}\left| \mathcal{K}(\basalfriction - \basalfriction_0)\right|^2\,dx,
\end{equation*}
is a Bilaplacian regularization term. Specifically, $\basalfriction_0$ is the constant function $\basalfriction_0(x)=10.5$, and $\mathcal{K}$ is the inverse of the solution operator for the following elliptic PDE:
\begin{align*}
-\gamma\Delta u +\delta\,u&=\regrhs\,\,\,\,\text{ in }\Omega,\\
\gamma \nabla u\cdot\normalvec+\regrobincoeff\,u&=0\,\,\,\text{ on }\partial\Omega.
\end{align*}
Recall that $\Omega$ is the 2D flattened version of the basal surface $\Gamma_b$. Here $\regrhs$ is a generic forcing term, and $\regrobincoeff=\sqrt{\delta \gamma}/1.42$ is a Robin boundary condition coefficient \cite{Roininen14}, and these quantities are different from $\bodyforce$ and $\stokesrobincoeff$ in the Stokes equation, \eqref{Stokeseqn}.
In all results except one, $\gamma$ is chosen so that the Morozov discrepancy principle is satisfied, i.e.,
\begin{equation*}
	\frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS = \frac{1}{2}\int_{\Gamma_{t}}||\xi||_{2}^{2}\,dS.
\end{equation*}
The value of $\gamma$ satisfying the Morozov discrepancy principle is $\gamma=7.3 \times 10^3$. The exception is Figure \ref{fig:gamma_sweep}, where we vary $\gamma$ to study how changing $\gamma$ impacts the effectiveness of our preconditioner. The constant $\delta$ is chosen so that the correlation length of functions drawn from the normal distribution with covariance $\mathcal{K}^{-2}$, given by $L=\sqrt{\gamma/\delta}$, is $1/10$th the radius of the domain. 
%, which is halfway between the maximum and minimum values of $\basalfriction_\text{true}$.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 impulse responses batches. These approximations are denoted by PSF (1), PSF (5), and PSF (25), respectively. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ often has negative values in its integral kernel when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $\bm{H}$. The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to a H-matrix approximation of the Hessian of the regularization term to form a H-matrix approximation of $\bm{H}_\text{gn}$. Straightforward calculation shows that the discretized Hessian of the regularization term is given by
$
(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
$
where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques. We factor the overall Gauss-Newton Hessian approximation, denoted $\bm{\preconditioner}$, using fast H-matrix methods, then and use the factorization as a preconditioner \cite{HLIBPRO}.

@@

Table \ref{tab:newton_convergence_table} shows the performance of our preconditioner for accelerating the solution of optimization problem \eqref{eq:deterministic_optimization_problem}. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and re-use it for all subsequent Gauss-Newton and Newton iterations. No preconditioning is used in the Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare our method with the most commonly used existing preconditioners: no preconditioning (NONE), and preconditioning by the regularization term in the Hessian (REG). In Figure \ref{fig:newton_iterates} we show select Newton iterates using our PSF (5) preconditioner. Using PSF (5) reduces the total number of Stokes PDE solves by a factor of six compared to no preconditioning, and by a factor of nine compared to regularization preconditioning\footnote{Interestingly, we observe that no preconditioning outperforms regularization preconditioning here because the noise level (and hence $\gamma$) is small.}.

@@

We presented an efficient matrix-free PSF method for approximating operators with locally supported non-negative integral kernels. The method only requires access to the operator via application of the operator to a small number of vectors. The idea of the method is to compute batches of impulse responses by applying the operator to Dirac combs of scattered point sources, then interpolate these impulse responses to approximate entries of the operator's integral kernel. The interpolation is based on a new principle we call ``local mean displacement invariance,'' which generalizes and improves upon classical local translation invariance. The ability to quickly approximate arbitrary integral kernel entries permits us to form a H-matrix approximation of the operator. Fast H-matrix arithmetic is then used to perform further linear algebra operations that cannot be performed easily with the original operator, such as matrix factorization and inversion. The supports of the impulse responses are estimated to be contained in ellipsoids, which are determined a-priori via a new procedure that involves applying the operator to a small number of polynomial functions. Point source locations for the impulse response batches are chosen using a greedy ellipsoid packing procedure, in which we choose as many impulse responses per batch as possible, while ensuring that the supports of the corresponding ellipsoids do not overlap. We applied the method to approximate the Gauss-Newton Hessian in a Stokes ice sheet inverse problem, and saw that the approximation substantially outperforms existing Hessian approximation methods. Although the method we presented is not applicable to all Hessians, it is applicable to many Hessians of practical interest. For these Hessians, our method offers a \emph{data-scalable} alternative to conventional low rank approximation, because our method can form high rank approximations of an operator using a small number of operator applies.

@@

\section{Ellipsoid intersection test}
\label{sec:fast_ellipsoid_intersection_test}
The procedure for choosing sample points relies on quickly determining whether two ellipsoids intersect. Let $E_1$ and $E_2$ be the ellipsoids defined as
\begin{align*}
	E_1 :=& \{x : (x - \spatialmean_1)^T \spatialcov_1^{-1} (x - \spatialmean_1) \le \tau^2\}, \\
	E_2 :=& \{x : (x - \spatialmean_2)^T \spatialcov_2^{-1} (x - \spatialmean_2) \le \tau^2\}, \\
\end{align*}
where $\spatialmean_1, \spatialmean_2 \in \mathbb{R}^\gdim$, and $\spatialcov_1, \spatialcov_2 \in \mathbb{R}^{\gdim \times \gdim}$ are positive definite. Let $K$ be the following one dimensional convex function:
\begin{equation*}
	K(s) := 1 - \frac{1}{\tau^2} (\spatialmean_1 - \spatialmean_2)^T \left(\frac{1}{1-s}\spatialcov_1 + \frac{1}{s}\spatialcov_2\right)^{-1}(\spatialmean_1 - \spatialmean_2)	
\end{equation*}
In \cite{GilitschenskiHanebeck12} it is shown that $E_1 \cap E_2 = \{\}$ if and only if $K(s) < 0$ for some $s\in (0,1)$.
% (see Figure \ref{fig:ellipsoid_intersection_test}).

>>>>

The ability to rapidly evaluate approximate kernel entries $\Aker(y,x)$ allows us to construct an H-matrix approximation, $\boldsymbol{\Aker}_H \approx \mathbf{\Aker}$, using the conventional adaptive cross H-matrix construction method (Section \ref{sec:h_matrix_construction_short}). In this method, one forms low rank approximations of off-diagonal blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. We then use H-matrix methods to convert $\boldsymbol{\Aker}_H$ into an H-matrix approximation $\mathbf{A}_H \approx \mathbf{A}$. Schematically the algorithm proceeds as follows:
\begin{equation*}
\Aop \longrightarrow \spatialvol, \spatialmean, \spatialcov \longrightarrow \underbrace{\{x_i\}_{i=1}^\nsamplepts}_{\text{as }S_b\text{'s}} \longrightarrow \underbrace{\{\impulseresponse_{x_i}\}_{i=1}^\nsamplepts}_{\text{as }\eta_b\text{'s}} \longrightarrow \boldsymbol{\Aker}_H \longrightarrow \mathbf{A}_H.
\end{equation*}
The complete algorithm for constructing $\mathbf{A}_H$ is shown in Algorithm \ref{alg:construct_Atilde}. The cost is discussed in Section \ref{sec:overall_cost}.

@@

The computational cost of our method may be divided into the costs to perform the following tasks: (1) Computing impulse response moments and batches; (2) Building the H-matrix; (3) Performing linear algebra operations with the H-matrix. This may optionally include modifying the H-matrix to make it symmetric positive semi-definite.
In practice, (1) is the dominant cost, because (1) is the only task that requires applying $\Aop$ to vectors, and our method is targeted at applications in which applying $\Aop$ to a vector requires an expensive computational procedure such as solving a large-scale PDE. All operations that do not require applications of $\Aop$ to vectors are nearly-linear, and therefore scalable, in the size of the problem, $N$.
We now describe these costs in detail. For convenience, Table \ref{tab:vars} lists variable symbols and their approximate sizes.

@@

\paragraph{(1) Computing impulse response moments and batches} Computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ requires applying $\Aop$ to $1$, $\gdim$, and $\gdim(\gdim+1)/2$ vectors, respectively.
Computing each impulse response batch requires applying $\Aop$ to one vector, so computing all impulse response batches $\{\eta_i\}_{i=1}^{\nbatch}$ requires $\nbatch$ operator applies. In total, computing all impulse response moments and batches therefore requires
\begin{equation*}
	1 + \gdim + \gdim(\gdim+1)/2 + \nbatch \qquad \text{operator applies.}
\end{equation*}
In a typical application one might have $\gdim=2$ and $\nbatch=5$, in which case a modest $11$ operator applies are required.

@@

\paragraph{(2) Building the H-matrix} The classical H-matrix construction technique
%, which is described in Appendix \ref{app:h_matrix}, 
requires evaluating $O(\hrank^2 \fedim \log \fedim)$ matrix entries of the approximation, where $\hrank$ is the H-matrix rank, i.e,  the maximum rank among the blocks of the H-matrix. To evaluate one matrix entry, first one must find the $\numnbr$ nearest sample points to a given point, where $\numnbr$ is the number of impulse responses used in the RBF interpolation. This is done using a precomputed kd-tree of sample points, and requires $O(\numnbr \log \nsamplepts)$ floating point and logical elementary operations. Second, one must find the mesh cells that the points $\{z_i\}_{i=1}^{\numnbr}$ reside. This is done using an AABB tree of mesh cells, and requires $O(\numnbr \log \fedim)$ elementary operations. Third, one must evaluate finite element basis functions on each of those cells, which requires $O(\numnbr)$ elementary operations. Finally, one must solve a $\numnbr \times \numnbr$ linear system to perform the radial basis function interpolation, which requires $O(\numnbr^3)$ elementary operations, which typically is negligible. Therefore, building the H-matrix requires
\begin{equation*}
	O\left(\left(\hrank^2 \fedim \log \fedim\right)\left(\numnbr \log \fedim + \numnbr^3\right)\right) \qquad \text{elementary operations}.
\end{equation*}
%This follows from the above discussion, and the fact that $\fedim \ge \nsamplepts$.

\paragraph{(3) Performing linear algebra operations with the H-matrix} It is well-known that H-matrix methods for matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion, and low rank updates require
\begin{equation}
\label{eq:H_matrix_cost_operations}
	O(\hrank^2 \fedim \log(\fedim)^a) \qquad \text{elementary operations},
\end{equation}
where $a \in \{0,1,2,3\}$ \cite{GrasedyckHackbusch03}. The precise cost depends on the type of H-matrix used and the operation being performed. In the numerical results (Section \ref{sec:numerical_results}), we use one matrix-matrix addition to add the H-matrix approximation of the data misfit term in the Hessian to the regularization term in the Hessian. Constructing a H-matrix representation of the regularization term in the Hessian requires one matrix inversion and two matrix-matrix multiplications. Then, we use one matrix factorization to form a preconditioner for the Newton linear system. 
%We perform low-rank updates to the H-matrix as the Newton iterations proceed, as per the discussion in Appendix \ref{sec:recycle_krylov}.
The rational method for modifying $\mathbf{A}_H$ to be positive semi-definite, described in Appendix \ref{sec:make_spd}, is performed using two matrix-matrix additions, $\ratord+1$ matrix-matrix multiplications, and one matrix inversion, where $2^\ratord$ is the rational function order.

@@

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a 2D surface embedded in 3D due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate. The parameter $\basalfriction$ is dually viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building our Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space.

@@

The reconstructed parameter field $\basalfriction$ is found as the solution to an optimization problem of the following form:
\begin{equation}
	\label{eq:deterministic_optimization_problem}
	\min_{\basalfriction} \quad J(\basalfriction) := J_d(\basalfriction) + J_r(\basalfriction).
\end{equation}
The first term in the objective function is
\begin{equation*}
	J_d(\basalfriction) := \frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS.
\end{equation*}
We call this term the data misfit, because it measures the difference between the observed data and the predicted data based on a candidate parameter $\basalfriction$. Here $\velocity = \velocity(\basalfriction)$ denotes the velocity field solving $\ref{Stokeseqn}$ for the given parameter $\basalfriction$. The second term,
\begin{equation*}
	J_r(\basalfriction) := \frac{1}{2}\int_{\Omega}\left| \mathcal{K}(\basalfriction - \basalfriction_0)\right|^2\,dx,
\end{equation*}
is a Bilaplacian regularization term. Specifically, $\basalfriction_0$ is the constant function $\basalfriction_0(x)=10.5$, and $\mathcal{K}$ is the inverse of the solution operator for the following elliptic PDE:
\begin{align*}
-\gamma\Delta u +\delta\,u&=\regrhs\,\,\,\,\text{ in }\Omega,\\
\gamma \nabla u\cdot\normalvec+\regrobincoeff\,u&=0\,\,\,\text{ on }\partial\Omega.
\end{align*}
Recall that $\Omega$ is the 2D flattened version of the basal surface $\Gamma_b$. Here $\regrhs$ is a generic forcing term, and $\regrobincoeff=\sqrt{\delta \gamma}/1.42$ is a Robin boundary condition coefficient \cite{Roininen14}, and these quantities are different from $\bodyforce$ and $\stokesrobincoeff$ in the Stokes equation, \eqref{Stokeseqn}.
In all results except one, $\gamma$ is chosen so that the Morozov discrepancy principle is satisfied, i.e.,
\begin{equation*}
	\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS = \int_{\Gamma_{t}}||\xi||_{2}^{2}\,dS.
\end{equation*}
The value of $\gamma$ satisfying the Morozov discrepancy principle is $\gamma=7.3 \times 10^3$. The exception is Figure \ref{fig:gamma_sweep}, where we vary $\gamma$ to study how changing $\gamma$ impacts the effectiveness of our preconditioner. The constant $\delta$ is chosen so that the correlation length of functions drawn from the normal distribution with covariance $\mathcal{K}^{-2}$, given by $L=\sqrt{\gamma/\delta}$, is $1/10$th the radius of the domain. 
%, which is halfway between the maximum and minimum values of $\basalfriction_\text{true}$.

@@

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 impulse responses batches. These approximations are denoted by PSF (1), PSF (5), and PSF (25), respectively. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel, especially when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $\bm{H}$. The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to a H-matrix approximation of the Hessian of the regularization term to form a H-matrix approximation of $\bm{H}_\text{gn}$. Straightforward calculation shows that the discretized Hessian of the regularization term is given by
$
(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
$
where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques. We factor the overall Gauss-Newton Hessian approximation, denoted $\bm{\preconditioner}$, using fast H-matrix methods \cite{RONKRIEMANNEMAIL}, then and use the factorization as a preconditioner \cite{HLIBPRO}.

@@

Table \ref{tab:newton_convergence_table} shows the performance of our preconditioner for accelerating the solution of optimization problem \eqref{eq:deterministic_optimization_problem}. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and re-use it for all subsequent Gauss-Newton and Newton iterations. No preconditioning is used in the Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare our method with the most commonly used existing preconditioners: no preconditioning (NONE), and preconditioning by the regularization term in the Hessian (REG). Using PSF (5) reduces the total number of Stokes PDE solves by a factor of six compared to no preconditioning, and by a factor of nine compared to regularization preconditioning\footnote{Interestingly, we observe that no preconditioning outperforms regularization preconditioning here because the noise level (and hence $\gamma$) is small.}. In Figure \ref{fig:newton_iterates} we show select Newton iterates using our PSF (5) preconditioner.

@@

We presented an efficient matrix-free PSF method for approximating operators with locally supported non-negative integral kernels. The method only requires access to the operator via application of the operator to a small number of vectors. The idea of the method is to compute batches of impulse responses by applying the operator to Dirac combs of scattered point sources, then interpolate these impulse responses to approximate entries of the operator's integral kernel. The interpolation is based on a new principle we call ``local mean displacement invariance,'' which generalizes and improves upon classical local translation invariance. The ability to quickly approximate arbitrary integral kernel entries permits us to form a H-matrix approximation of the operator. Fast H-matrix arithmetic is then used to perform further linear algebra operations that cannot be performed easily with the original operator, such as matrix factorization and inversion. The supports of the impulse responses are estimated to be contained in ellipsoids, which are determined a-priori via a new procedure that involves applying the operator to a small number of polynomial functions. Point source locations for the impulse response batches are chosen using a greedy ellipsoid packing procedure, in which we choose as many impulse responses per batch as possible, while ensuring that the corresponding ellipsoids do not overlap. We applied the method to approximate the Gauss-Newton Hessian in a Stokes ice sheet inverse problem, and saw that the approximation substantially outperforms existing Hessian approximation methods. Although the method we presented is not applicable to all Hessians, it is applicable to many Hessians of practical interest. For these Hessians, our method offers a \emph{data-scalable} alternative to conventional low rank approximation, because our method can form high rank approximations of an operator using a small number of operator applies.

@@

\section{Ellipsoid intersection test}
\label{sec:fast_ellipsoid_intersection_test}
The procedure for choosing sample points relies on quickly determining whether two ellipsoids intersect. Let $E_1$ and $E_2$ be the ellipsoids defined as
\begin{align*}
	E_1 :=& \{x : (x - \spatialmean_1)^T \spatialcov_1^{-1} (x - \spatialmean_1) \le \tau^2\}, \\
	E_2 :=& \{x : (x - \spatialmean_2)^T \spatialcov_2^{-1} (x - \spatialmean_2) \le \tau^2\},
\end{align*}
where $\spatialmean_1, \spatialmean_2 \in \mathbb{R}^\gdim$, and $\spatialcov_1, \spatialcov_2 \in \mathbb{R}^{\gdim \times \gdim}$ are positive definite. Let $K$ be the following one dimensional convex function:
\begin{equation*}
	K(s) := 1 - \frac{1}{\tau^2} (\spatialmean_1 - \spatialmean_2)^T \left(\frac{1}{1-s}\spatialcov_1 + \frac{1}{s}\spatialcov_2\right)^{-1}(\spatialmean_1 - \spatialmean_2).
\end{equation*}
In \cite{GilitschenskiHanebeck12} it is shown that $E_1 \cap E_2 = \{\}$ if and only if $K(s) < 0$ for some $s\in (0,1)$.
% (see Figure \ref{fig:ellipsoid_intersection_test}).
